### 2016-07-31 Projects/DelCo_2012/Report/DelCo_2012.tex
Added Jack's edits and Changed CAFS to CAF in gas cooling captions

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires and also examined using them for structural fire fighting. Other departments across the country began to experiment with CAFS for use in structural fire fighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable. Yet, after decades of fire department trials, CAF has still not been widely incorporated into structural fire fighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural fire fighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. wood-based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets as well as carpeting and upholstered furniture.

@@

The fire suppression series included 20 experiments in the single-story experimental structures and one in the two-story structure. Twelve experiments were conducted in the single-story structures (6 in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied and included straight stream, solid stream, and narrow fog stream. The two-story test used both water and CAF from a straight stream nozzle.

@@

Each structure had burn room, approximately 6.0~m (19.6~ft) by 4.4~m (14.3~ft) and 2.74~m (9.0~ft) in height and a hallway connecting the burn room to the front of the structure via a small entry foyer. The hallway was 0.95~m (3.1~ft) wide and 5.4~m (17.6~ft) long and the entry foyer was 1.2~m (4.1~ft) by 1.9~m (6.1~ft). The ceiling height in the hallway and entry foyer was 2.4~m (8~ft).  The open door on north side from the exterior to the entry foyer was 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide. The opening on the south face of the structure from the exterior to the burn room was 2.4~m (7.75~ft) high and 1.09~m (3.6~ft) wide. A dimensioned, plan view of the structure is included in Figure~\ref{fig:Test_Structure_Floor_Plan}.

@@

The structures were instrumented for temperature, gas velocity, and heat flux measurements. Gas temperatures in the burn rooms were measured with bare-bead, Chromel-Alumel (type K) thermocouples. Additional single, sheathed thermocouples were installed in conjunction with the bi-directional probes for gas velocity measurements. The single thermocouples were bare-bead, Chromel-Alumel (type K) thermocouples with a 1.0~mm (0.04~in) nominal diameter. The thermocouple was protected with an 3.2~mm (0.125~in) diameter inconel sheath. The sheathing protects the thermocouple from water but slows the response time because of the additional thermal mass. Schmidt-Boelter gauges were used to measure both total heat flux and radiant heat flux (radiometer). A radiometer is a total heat flux gauge with a zirconium plate to prevent contributions from convective heat transfer. A legend, which clarifies the instrumentation schematics discussed in the follow sections, is included in Figure~\ref{fig:Instrumentation_Legend}.

@@

The gas cooling tests included 4 bare-bead thermocouple arrays  a bi-directional probe plus solid thermocouple array, and a total heat flux gauge/radiometer pair. Figure~\ref{fig:Gas_Cooling_Instrumentation_Dimensions} provides the positions within the burn room and gas cooling room where the senors were located (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each of the bare-bead thermocouple array had 11 thermocouples. The top thermocouple was placed at the ceiling and each subsequent thermocouple was spaced 0.3~m (1~ft) apart with the bottom thermocouple being 3.05~m (10ft) below the ceiling. There were 6 velocity probes and solid thermocouples centered at the external doorway to the gas cooling room. The top probes were 0.15~m (0.5~ft) below the door soffit, the second probes were 0.3~m (1~ft), and the remaining 4 were spaced 0.3~m (1~ft) apart with the bottom probe being 1.52~m (5~ft) below the soffit. The total heat flux gauge/radiometer were set to be 0.15~m (0.5~ft) off the ground and aimed to view the ceiling.

@@

The fire suppression testing in the single-story structure included 2 bare-bead thermocouple arrays, three bi-directional probe plus solid thermocouple arrays, 2 total heat flux gauge pairs, and 2 total heat flux gauge/radiometer pairs. Figure~\ref{fig:Fire_Suppression_Instrumentation_Dimensions} provides the positions within the burn room, hallway, and entrance foyer where the senors were located (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each bare-bead thermocouple array had 8 thermocouples. The top sensor was 0.03~m (1~in) below the ceiling and the remaining thermocouples were space 0.3~m (1~ft) apart with the bottom thermocouple being 2.13~m (7~ft) below the ceiling. The three bi-directional probe arrays had unique sensor locations. For the window array there were four sensor pairs located 0.28~m (0.95~ft), 0.58~m (1.90~ft), 0.84~m (2.85~ft), and 1.16~m (3.8~ft) below the soffit. For the hallway array, the 7 sensor pairs started 0.3~m (1~ft) below the ceiling and were spaced every 0.3~m (1~ft) ending 2.13~m (7~ft) below the ceiling. The doorway array also had 7 sensor pairs, but the first sensors were located 0.15~m (0.5~ft) below the soffit. The remaining sensors were spaced every 0.3~m (1~ft) with the bottom pair 1.83~m (6~ft) below the doorway soffit. The total heat flux gauge/radiometer pairs were set to be 0.15~m (0.5~ft) off the ground and aimed to view the ceiling. The pairs of total heat flux gauges were set 1~m (3~ft) off the ground with one sense ``looking'' vertical and the other horizontal.

@@

Fire suppression testing in the two-story structure included similar types instrumentation but included more sensors. Three bare-bead thermocouple arrays and 2 bi-directional probe plus solid thermocouples arrays were used in the first floor. Figure~\ref{fig:fire_supp_first_2story} shows the positions of the instrumentation (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each of the bare-bead arrays had 8 thermocouples with the same spacing as the single-story tests. The bi-directional probe and solid thermocouple arrays also had 8 sensors with the first probe 0.08~m (0.25~ft) below the soffit with the remaining probes at 0.34~m (1.1~ft), 0.61~m (2~ft), 0.88~m (2.9~ft), 1.15~m (3.7~ft), 1.42~m (4.7~ft), 1.68~m (5.5~ft), and 1.95~m (6.4~ft) below the soffit.

@@

Bi-directional probes and single thermocouples were used to measure gas velocity. The bi-directional probes used similar pressure transducers as those used for the differential pressure measurements discussed above. A bare-bead Type K thermocouple was co-located with each probe. A gas velocity measurement study examining the doorway flow of pre-flashover compartment fires yielded expanded uncertainty measurements ranging from $\pm$ 0.14 to $\pm$ 0.22 for bi-directional probes of similar design~\cite{Bryant:FSJ2009}. The total expanded uncertainty for gas velocity in these experiments is estimated to be $\pm$ 18~\%.

@@

In June 2015, two stacks of pallets were added to the fuel load consisting of three sofas; OSB lining on the ceiling and walls; and the polypropylene carpeting over puf padding. Each stack had 10 pallets and an approximate mass of 175~kg (386~lbs). Hay from bales averaging a mass of 12~kg (26~lbs) was added for ignition purposes.

@@

After weighing each of the bins, the mass of water (kg) in each bin was plotted by position to visualize the spray distribution. Figure~\ref{fig:Burn_Building_Test_1} shows that with a straight stream in the back position, the majority of the water accumulated along the north and east walls of the gas cooling room within the burn building (see Figure~\ref{fig:Delaware_County,_PA_Burn_Building_Layout}). Figure~\ref{fig:Burn_Building_Test_7} shows the difference of a fog stream, as the distribution of water is far more dispersed around the gas cooling room. The remainder of the burn building spray density figures are included in Appendix~\ref{app:spray_density}.

@@

To better quantify the effectiveness of the suppression agents (CAF and water), the maximum temperature differences associated with each gas cooling event for each agent from the three thermocouple trees were analyzed. Figure~\ref{fig:combined_all} compares the temperature differences at each thermocouple position (via a scatterplot) for 30 of the 88 experiments. The 30 experiments are actually pairs of experiments, 15 water and 15 CAF, which were directly comparable (see Table~\ref{tab:Gas_Cooling_Descriptions}). These experiments included solid stream, straight streams, and narrow fog streams at both nozzle positions. While the total number of experiments was 88, there were a number of experiments that did not have a directly comparable experiment between water and CAF. For example, a set of 12 Class A solution experiments were conducted. Gas cooling experiments were conducted with water applied in a wide fog stream. Since no significant differences were noted compared to the narrow fog stream, a comparable series with a wide fog CAF was not conducted. Similarly, larger water flows were examined, but 120 gpm was the maximum water flow that could be generated with the CAFS, so those experiments are also not included in this comparison. The Mitchell report includes the data for all 88 experiments \cite{Mitchell:1}.

@@

The data can be separated to examine the impact of nozzle position (back or middle position) on the ability to cool gases for each of the suppression agents. Figures~\ref{fig:CAFS_Water_mid} and \ref{fig:CAFS_Water_full} show the cooling comparison for 12 pairs of experiments in the back position and 3 pairs of experiments in the middle position respectively.

@@

For the 12 middle position replicate comparisons (Figure~\ref{fig:CAFS_Water_mid}), the results (via the dashed red line) show that water has a slight advantage in gas cooling relative to CAF. At the back position (Figure~\ref{fig:CAFS_Water_full}), for the 3 replicate comparison, the least squared error line falls within the experimental uncertainty. This indicates an inability to distinguish between the two suppression agents with statistical confidence.

@@

There is a difference in the amount of direct impact from the suppression agent at the three spatial locations. Therefore, the maximum temperature differences for each thermocouple on each of the three arrays should also be analyzed separately. Figures~\ref{fig:CAFS_Water_A1_all}, \ref{fig:CAFS_Water_A2_all}, and \ref{fig:CAFS_Water_A3_all} show the temperature cooling scatter for thermocouple trees A1, A2, and A3 respectively. Note that thermocouple array A3, the array which has the least amount of direct impact from the suppression agents, has the closest fit compared to the solid black line that represents uniform impact for both suppression agents. The fit lines for A1 and A3 also fall within the experimental uncertainty, meaning that differences in the impact of CAF and water are indistinguishable. Thermocouple tree A2, which gets the most significant direct impact from the suppression agents, shows a bias towards water being more effective than CAF on cooling the thermocouples.

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A1 Experimental Pairs}
	\label{fig:CAFS_Water_A1_all}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A2 Experimental Pairs}
	\label{fig:CAFS_Water_A2_all}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A3 Experimental Pairs}
	\label{fig:CAFS_Water_A3_all}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A1 Replicates}
	\label{fig:CAFS_Water_A1_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A2 Replicates}
	\label{fig:CAFS_Water_A2_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A3 Replicates}
	\label{fig:CAFS_Water_A3_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A1 Replicates}
	\label{fig:CAFS_Water_A1_back}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A2 Replicates}
	\label{fig:CAFS_Water_A2_back}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A3 Replicates}
	\label{fig:CAFS_Water_A3_back}
\end{figure}

@@

There were 12 suppression tests conducted in the single-story purpose-built concrete structure. Eight of the experiments used a fuel load comprised of typical household furnishings that included wooden wall paneling, two sofas, two overstuffed chairs, carpet, and padding (Section~\ref{sec:fire_suppression_furniture_fuel}). The other four experiments used wood pallet fuel load in a space with the ceiling and walls composed of wood-based materials (Section~\ref{sec:fire_suppression_pallet_fuel}). Ignition occurred in the rear of the single-story structures and the available ventilation was through the front doorway and for eight of the experiments a rear window was ``opened''to provide additional ventilation (see Table~\ref{tab:Test_Descriptions}). The fires were intended to replicate post-flashover residential room and contents fires. Hose lines were pre-deployed with fixed nozzles set in two locations: one in the hallway on the approach to the burn room (hallway nozzle) and one just inside the burn room to the right side of the hallway (door nozzle). Each nozzle flowed 15~s of water or compressed air foam depending on the test. The hallway nozzle was intended to simulate a crew advancing into a structure towards the burn room. Suppression from this location would be in the form of gas cooling as the stream was not aimed at the seat of the fire (indirect interior attack).  After the hallway nozzle was shutdown, there was a 15~s time period to simulate the crew advancing into the doorway of the fire room. The door nozzle was positioned just to the side of the entry door.  It was also opened and flowed for approximately 15~s. The door nozzles was aimed in the direction of the seat of the fire (direct interior attack). Table~\ref{tab:Test_Descriptions} provides an overview of the suppression tests conducted.

@@

Note that while there are some differences in the changes in temperature and heat flux when comparing CAF to Water, differences are not substantial. Both suppression agents had similar, significant impact in reducing the high-temperature, high-heat flux environment that existed before suppression. The first (hallway) suppression, an indirect attack, did a majority of the cooling in all of the tests. This illustrates the impact of gas cooling, independent of suppression agent. In several cases rekindling occurred, as seen by the temperature increase between the two nozzle applications, but all cases the second (room) attack controlled the fire.

@@

There were 7 suppression tests conducted in the two-story purpose-built structure. All seven of the tests used a fuel load consisting of three sofas, polypropylene carpeting over polyurethane foam padding, and OSB paneling along three walls and the ceiling, and five of the seven tests had two stacks of wooden pallets in addition to the initial fuel load (Section~\ref{sec:fire_suppression_furniture_fuel}). Ignition occurred in the southwest corner of the basement and available ventilation was through the basement double doors and the 1st floor door on the south side of the structure. The fires were intended to replicate room and contents basement fires typically seen within residential structures. Suppression occurred at the basement double doors of the structure. Table~\ref{tab:Test_Descriptions_2} provides an overview of the suppression tests conducted.

@@

This study focused on two key types of fire experiments; gas cooling and fire suppression. In each type of experiment a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates. In both types of experiments there were no significant measurable differences between the use of plain water or compressed air foam. To expand upon the 2012 and 2013 tests described in~\cite{Mitchell:1}, additional fire suppression experiments were conducted. One set of experiments used only wood-based fuels, to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. A final experimental series, used a larger fire room, with more ventilation. In this final series, another change was made, the nozzle was hand held and could be moved as needed to effectively extinguish the fire. This section will discuss the overall results of the study and provide potential explanations to support them.

@@

Thank you for assisting NIST in generating data that will assist the fire service in making tactical choices.

>>>>

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires and also considered their use for structural fire fighting. Other departments across the country began to experiment with CAFS for use in structural fire fighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable. Yet, after decades of fire department trials, CAF has still not been widely incorporated into structural fire fighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural fire fighting were invited and present at many of the fire suppression experiments. The experiments were planned during the workshop.  Three types of experiments examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. wood-based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets, as well as carpeting and upholstered furniture.

@@

The fire suppression series included 20 experiments. Twelve experiments were conducted in the single-story structures (6 in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied and included straight stream, solid stream, and narrow fog stream. The six experiments in the two-story structure used both water and CAF from a straight stream nozzle.  In addition two CAF experiments were conducted in the two-story with a solid stream.

@@

Each structure had a burn room, approximately 6.0~m (19.6~ft) by 4.4~m (14.3~ft) and 2.74~m (9.0~ft) in height and a hallway connecting the burn room to the front of the structure via a small entry foyer. The hallway was 0.95~m (3.1~ft) wide and 5.4~m (17.6~ft) long and the entry foyer was 1.2~m (4.1~ft) by 1.9~m (6.1~ft). The ceiling height in the hallway and entry foyer was 2.4~m (8~ft).  The open door on north side from the exterior to the entry foyer was 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide. The opening on the south face of the structure from the exterior to the burn room was 2.4~m (7.75~ft) high and 1.09~m (3.6~ft) wide. A dimensioned, plan view of the structure is included in Figure~\ref{fig:Test_Structure_Floor_Plan}.

@@

The structures were instrumented for temperature, gas velocity, and heat flux measurements. Gas temperatures in the burn rooms were measured with bare-bead, Chromel-Alumel (type K) thermocouples. Additional single, sheathed thermocouples were installed in conjunction with the bi-directional probes for gas velocity measurements. The single thermocouples were bare-bead, Chromel-Alumel (type K) thermocouples with a 1.0~mm (0.04~in) nominal diameter. The thermocouple was protected with an 3.2~mm (0.125~in) diameter inconel sheath. The sheathing protects the thermocouple from water but slows the response time because of the additional thermal mass. Schmidt-Boelter gauges were used to measure both total heat flux and radiant heat flux (radiometer). A radiometer is a total heat flux gauge with a zirconium plate to prevent contributions from convective heat transfer. A legend, which clarifies the instrumentation schematics discussed in the following sections, is included in Figure~\ref{fig:Instrumentation_Legend}.

@@

The gas cooling tests included 4 bare-bead thermocouple arrays, a bi-directional probe with a thermocouple array, and a total heat flux gauge/radiometer pair. Figure~\ref{fig:Gas_Cooling_Instrumentation_Dimensions} provides the positions within the burn room and gas cooling room where the senors were located (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each of the bare-bead thermocouple arrays had 11 thermocouples. The top thermocouple was placed at the ceiling and each subsequent thermocouple was spaced 0.3~m (1~ft) apart with the bottom thermocouple being 3.05~m (10ft) below the ceiling. There were 6 velocity probes and solid thermocouples centered at the external doorway to the gas cooling room. The top probes were 0.15~m (0.5~ft) below the door soffit, the second probes were 0.3~m (1~ft), and the remaining 4 were spaced 0.3~m (1~ft) apart with the bottom probe being 1.52~m (5~ft) below the soffit. The total heat flux gauge/radiometer set were positioned 0.15~m (0.5~ft) off the ground and aimed to view the ceiling.

@@

The fire suppression testing in the single-story structure included 2 bare-bead thermocouple arrays, three bi-directional probe plus solid thermocouple arrays, 2 total heat flux gauge pairs, and 2 total heat flux gauge/radiometer pairs. Figure~\ref{fig:Fire_Suppression_Instrumentation_Dimensions} provides the positions within the burn room, hallway, and entrance foyer where the senors were located (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each bare-bead thermocouple array had 8 thermocouples. The top sensor was 0.03~m (1~in) below the ceiling and the remaining thermocouples were spaced 0.3~m (1~ft) apart with the bottom thermocouple being 2.13~m (7~ft) below the ceiling. The three bi-directional probe arrays had unique sensor locations. For the window array, there were four sensor pairs located 0.28~m (0.95~ft), 0.58~m (1.90~ft), 0.84~m (2.85~ft), and 1.16~m (3.8~ft) below the soffit. For the hallway array, the 7 sensor pairs started 0.3~m (1~ft) below the ceiling and were spaced every 0.3~m (1~ft) ending 2.13~m (7~ft) below the ceiling. The doorway array also had 7 sensor pairs, but the first sensors were located 0.15~m (0.5~ft) below the soffit. The remaining sensors were spaced every 0.3~m (1~ft) with the bottom pair 1.83~m (6~ft) below the doorway soffit. The total heat flux gauge/radiometer pairs were set to be 0.15~m (0.5~ft) off the ground and aimed to view the ceiling. The pairs of total heat flux gauges were set 1~m (3~ft) off the ground with one sensor ``looking'' vertical and the other horizontal.

@@

Fire suppression testing in the two-story structure included similar types of instrumentation but included more sensors. Three bare-bead thermocouple arrays and 2 bi-directional probe and thermocouples arrays were used in the first floor. Figure~\ref{fig:fire_supp_first_2story} shows the positions of the instrumentation (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each of the bare-bead arrays had 8 thermocouples with the same spacing as the single-story tests. The bi-directional probe and solid thermocouple arrays also had 8 sensors with the first probe 0.08~m (0.25~ft) below the soffit with the remaining probes at 0.34~m (1.1~ft), 0.61~m (2~ft), 0.88~m (2.9~ft), 1.15~m (3.7~ft), 1.42~m (4.7~ft), 1.68~m (5.5~ft), and 1.95~m (6.4~ft) below the soffit.

@@

Bi-directional probes and single thermocouples were used to measure gas velocity. The bi-directional probes used similar pressure transducers as those used for the differential pressure measurements discussed above. Bare-bead Type K thermocouples are co-located with each probe. A gas velocity measurement study examining the doorway flow of pre-flashover compartment fires yielded expanded uncertainty measurements ranging from $\pm$ 0.14 to $\pm$ 0.22 for bi-directional probes of similar design~\cite{Bryant:FSJ2009}. The total expanded uncertainty for gas velocity in these experiments is estimated to be $\pm$ 18~\%.

@@

In June 2015, two stacks of pallets were added to the fuel load consisting of three sofas; OSB lining on the ceiling and walls; and the polypropylene carpeting over polyurethane foam padding. Each stack had 10 pallets and an approximate mass of 175~kg (386~lbs). Hay from bales averaging a mass of 12~kg (26~lbs) was added for ignition purposes.

@@

After weighing each of the bins, the mass of water (kg) in each bin was plotted by position to visualize the spray distribution. Figure~\ref{fig:Burn_Building_Test_1} shows that with a straight stream in the back position, the majority of the water accumulated along the north and east walls of the gas cooling room within the burn building (see Figure~\ref{fig:Delaware_County,_PA_Burn_Building_Layout}). Figure~\ref{fig:Burn_Building_Test_7} shows the results from a fog stream, where the distribution of water is far more dispersed around the gas cooling room. The remainder of the burn building spray density figures are included in Appendix~\ref{app:spray_density}.

@@

To better quantify the effectiveness of the suppression agents (CAF and water), the maximum temperature differences associated with each gas cooling event for each agent from the three thermocouple trees were analyzed. Figure~\ref{fig:combined_all} compares the temperature differences at each thermocouple position (via a scatterplot) for 30 of the 88 experiments. The 30 experiments are actually pairs of experiments, 15 water and 15 CAF, which were directly comparable (see Table~\ref{tab:Gas_Cooling_Descriptions}). These experiments included solid stream, straight streams, and narrow fog streams at both nozzle positions. While the total number of experiments was 88, there were a number of experiments that did not have a directly comparable experiment between water and CAF. For example, a set of 12 Class A solution experiments were conducted. Gas cooling experiments were also conducted with water applied in a wide fog stream. Since no significant differences were noted compared to the narrow fog stream experiments, a comparable series with a wide fog CAF was not conducted. Similarly, larger water flows were examined, but 120 gpm was the maximum water flow that could be generated with the CAFS, so those experiments are also not included in this comparison. The Mitchell report includes the data for all 88 experiments \cite{Mitchell:1}.

@@

The data can be separated to examine the impact of nozzle position (back or middle position) on the ability to cool gases for each of the suppression agents. Figures~\ref{fig:CAFS_Water_mid} and \ref{fig:CAFS_Water_full} show the cooling comparison for 12 pairs of experiments in the back position and 3 pairs of experiments in the middle position, respectively.

@@

For the 12 middle position replicate comparisons (Figure~\ref{fig:CAFS_Water_mid}), the results (via the dashed red line) show that water has a slight advantage in gas cooling relative to CAF. At the back position (Figure~\ref{fig:CAFS_Water_full}), for the 3 replicate comparisons, the least squared error line falls within the experimental uncertainty. This indicates an inability to distinguish between the two suppression agents with statistical confidence.

@@

There is a difference in the amount of direct impact from the suppression agent at the three spatial locations. Therefore, the maximum temperature differences for each thermocouple on each of the three arrays should also be analyzed separately. Figures~\ref{fig:CAFS_Water_A1_all}, \ref{fig:CAFS_Water_A2_all}, and \ref{fig:CAFS_Water_A3_all} show the temperature cooling scatter for thermocouple trees A1, A2, and A3, respectively. Note that thermocouple array A3, the array which has the least amount of direct impact from the suppression agents, has the closest fit compared to the solid black line that represents uniform impact for both suppression agents. The fit lines for A1 and A3 also fall within the experimental uncertainty, meaning that differences in the impact of CAF and water are indistinguishable. Thermocouple tree A2, which gets the most significant direct impact from the suppression agents, shows a bias towards water being more effective than CAF on cooling the thermocouples.

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A1_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for All Thermocouple Tree A1 Experimental Pairs}
	\label{fig:CAFS_Water_A1_all}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A2_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for All Thermocouple Tree A2 Experimental Pairs}
	\label{fig:CAFS_Water_A2_all}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A3_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for All Thermocouple Tree A3 Experimental Pairs}
	\label{fig:CAFS_Water_A3_all}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A1_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A1 Replicates}
	\label{fig:CAFS_Water_A1_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A2_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A2 Replicates}
	\label{fig:CAFS_Water_A2_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A3_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A3 Replicates}
	\label{fig:CAFS_Water_A3_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A1_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A1 Replicates}
	\label{fig:CAFS_Water_A1_back}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A2_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A2 Replicates}
	\label{fig:CAFS_Water_A2_back}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A3_scatter}
	\caption{CAF vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A3 Replicates}
	\label{fig:CAFS_Water_A3_back}
\end{figure}

@@

There were 12 suppression tests conducted in the single-story purpose-built concrete structure. Eight of the experiments used a fuel load comprised of typical household furnishings that included wooden wall paneling, two sofas, two overstuffed chairs, carpet, and padding (Section~\ref{sec:fire_suppression_furniture_fuel}). The other four experiments used wood pallet fuel load in a space with the ceiling and walls composed of wood-based materials (Section~\ref{sec:fire_suppression_pallet_fuel}). Ignition occurred in the rear of the single-story structures and the available ventilation was through the front doorway and for eight of the experiments a rear window was ``opened''to provide additional ventilation (see Table~\ref{tab:Test_Descriptions}). The fires were intended to replicate post-flashover residential room and contents fires. Hose lines were pre-deployed with fixed nozzles set in two locations: one in the hallway on the approach to the burn room (hallway nozzle) and one just inside the burn room to the right side of the hallway (door nozzle). Each nozzle flowed 15~s of water or compressed air foam depending on the test. The hallway nozzle was intended to simulate a crew advancing into a structure towards the burn room. Suppression from this location would be in the form of gas cooling as the stream was not aimed at the seat of the fire (indirect interior attack).  After the hallway nozzle was shutdown, there was a 15~s time period to simulate the crew advancing into the doorway of the fire room. The door nozzle was positioned just to the side of the entry door.  It was also opened and flowed for approximately 15~s. The door nozzle was aimed in the direction of the seat of the fire (direct interior attack). Table~\ref{tab:Test_Descriptions} provides an overview of the suppression tests conducted.

@@

Note that while there are some differences in the changes in temperature and heat flux when comparing CAF to Water, the differences are not substantial. Both suppression agents had similar, significant impacts in reducing the high-temperature, high-heat flux environment that existed before suppression. The first (hallway) suppression, an indirect attack, did a majority of the cooling in all of the tests. This illustrates the impact of gas cooling, independent of suppression agent. In several cases rekindling occurred, as seen by the temperature increase between the two nozzle applications, but all cases the second (room) attack controlled the fire.  However in most cases rekindling occured as nozzle movement was needed to completely extinguish the fire.

@@

There were 7 suppression tests conducted in the two-story purpose-built structure. All seven of the tests used a fuel load consisting of three sofas, polypropylene carpeting over polyurethane foam padding, and OSB paneling along three walls and the ceiling.  Five of the seven tests also had two stacks of wooden pallets in addition to the initial fuel load (Section~\ref{sec:fire_suppression_furniture_fuel}). Ignition occurred in the southwest corner of the basement and available ventilation was through the basement double doors and the 1st floor door on the south side of the structure. The fires were intended to replicate room and contents basement fires typically seen within residential structures. Suppression occurred at the basement double doors of the structure. Table~\ref{tab:Test_Descriptions_2} provides an overview of the suppression tests conducted.

@@

This study focused on two key types of fire experiments; gas cooling and fire suppression. In each type of experiment a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates. In both types of experiments there were no significant measurable differences between the use of plain water or compressed air foam. To expand upon the 2012 and 2013 tests described in~\cite{Mitchell:1}, additional fire suppression experiments were conducted. One set of experiments used only wood-based fuels to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. A final experimental series used a larger fire room with more ventilation. In this final series, another change was made: the nozzle was hand held, and could be moved as needed to effectively extinguish the fire. This section will discuss the overall results of the study and provide potential explanations to support them.

@@

Thanks to all involved for assisting in generating data that will aid the fire service in making evidence based tactical choices.
